---
title: "ZebraFinchHeatDeath(Exp1)_RNAseqAnalyses"
output: html_notebook
---

#DESeq2 Analyses

#### Resources and Citations:
Love et al 2016 DESeq2 GenomeBiology
 https://bioconductor.riken.jp/packages/2.14/bioc/vignettes/DESeq2/inst/doc/DESeq2.pdf
http://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html

#### Install the DESeq2 package if you have not already

---
  #try http:// if https:// URLs are not supported
  # if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
 BiocManager::install("DESeq2")

---

## Load the DESeq2 library 
library(DESeq2)

  ##### If needed, use the Session menu to set working directory To Source File Directory

# Dataset:Deduplicated Pipeline to obtain read counts for a complete dataset of blood RNAseq sameples.
## Individuals used in this analyses: Baseline blood samples, from pubescent females (N=17), that either succumbed (n=10), or survived (n=7) to an acute 5 hour heat stress.
---
##  1.1 Input data   ##############
  ## Input the count data, the gene(/transcript) count matrix and labels
  ## this starts from with output from PrepDE.py from Ballgown folder.
```{r}
countdata.1 <- as.matrix(read.csv("gene_count_matrix_Dedup_BaselineHeatDeath_gene_ids_OnlyPubescent.csv", row.names="gene_id"))
dim(countdata.1)
head(countdata.1)
```

## Input the meta data or phenotype data
# Note: The PHENO_DATA file contains information on each sample including whether they Succumb to the heat stres (Yes or No), and that this is paired-end RNAseq data.
```{r}
coldata.1 <-(read.table("PHENO_DATA_Baseline_DeDupHeatDeath_OnlyPubescent_n19.txt", header=TRUE, row.names=1))
dim(coldata.1)
head(coldata.1)
```
### Check all sample IDs in colData are also in CountData and match their orders
```{r}
all(rownames(coldata.1) %in% colnames(countdata.1))
counts.1 <- countdata.1[, rownames(coldata.1)]
all(rownames(coldata.1) == colnames(counts.1))
```

## Create the DESEQ dataset and define the statistical model (page 6 of the manual)
```{r}
dds_1 <- DESeqDataSetFromMatrix(countData = countdata.1, colData=coldata.1,  design = ~succumb)
#look at it
dds_1
```

### Prefiltering - filter lowly expressed genes, which have a reduced ability to detect differences among groups, and would increase the number of multiple testing.
  ## Manual - starting at  1.3.6 
  ## Here we perform a minimal pre-filtering to remove rows (genes) that have less than, on average across individuals, 100 reads mapped per individual.  17 individuals x 100 reads = 1700 reads. A gene must have at least 1700 reads total to be included for analysis.
```{r}
dds_1f <- dds_1[ rowSums(counts(dds_1)) > 1700, ]
# How many genes are left after filtering the lowest expressed genes?
dds_1f
```

  # Total annotated genes in genome assembly: 23511
  # filter genes with average 100 reads per sample (1700 reads) = 9648  genes
  
## Set factors for statistical analyses
  ## Succumb column in the PHENO_DATA: No is the control (survived), Yes is succumb, i.e. the "treatment group"
# example:
#dds$condition <- factor(dds$condition, levels=c("untreated","treated"))
```{r}
dds_1f$condition <- factor(dds_1f$succumb, levels=c("No","Yes"))
```

#### 1.4 Differential expression analysis
```{r}
dds_1f <- DESeq(dds_1f)
res <- results(dds_1f)
res
```

###  Evaluate results
```{r}
  # We can order our results table by the smallest adjusted p value:
  resOrdered <- res[order(res$pvalue),]
  resOrdered
```
  # Summarize some basic tallies using the summary function the default is p<0.1.
```{r}
  summary(res)
```

# How many DEG at adjusted p-values < 0.05?
```{r}
  sum(res$padj < 0.05, na.rm=TRUE)
```

#If the adjusted p value will be a value other than 0.1, alpha should be set to that value:
## Here set to 0.05  
```{r}
  res05 <- results(dds_1f, alpha=0.05)
  summary(res05)
  sum(res05$padj < 0.05, na.rm=TRUE)
```

## write out the results file
```{r}
write.csv(as.data.frame(resOrdered),file="Submit_DGESeq_results_Dataset2-DeDup_Females_OnlyPubescent.csv") 
```
  
#####################  
###    1.5.1 MA-plot
  ##plotMA shows the log2 fold changes attributable to a given variable over the meanof normalized counts. 
  ## Points will be colored red if the adjusted p value is less than 0.1. 
  ## Points which fall out of the window are plotted as open triangles pointing either up or down
```{r}
plotMA(res, main="DESeq2", ylim=c(-8,8))
  
  #After calling plotMA, one can use the function identify to interactively detect the row number of individual genes by clicking on the plot. 
  # One can then recover the gene identiers by saving the resulting indices:
idx <- identify(res$baseMean, res$log2FoldChange)
    # after selecting a gene. You need to press escape to move on
  rownames(res)[idx]
```
    
##  1.5.2 Plot counts - sanity check!
  # You can select the gene to plot by rowname or by numeric index.
```{r}
## This gene has a negative logfold change, so should be downregulated in the succumb group
plotCounts(dds_1f, gene="THOC6", intgroup="succumb")
```

```{r}
## You can plot the gene with th lowest adjusted P-value.
## This gene has a postive logfold change so should be upregulated 
plotCounts(dds_1f, gene=which.min(res$padj), intgroup="succumb")
```
### 2.1.2 Extracting transformed values
```{r}
  rld <- rlog(dds_1f)
  vsd <- varianceStabilizingTransformation(dds_1f)
  head(assay(rld), 3)
``` 


####  Make a heatmap of the top differentially expressed genes adjusted p-value < 0.1
```{r}
library("pheatmap")
# Extract VSD matrix
vsd_mat <- assay(vsd)

# Order by p-value
res_ordered <- res[order(res$pvalue), ]

## Keep genes with adjusted p-value < 0.1
res_sig <- res[!is.na(res$padj) & res$padj < 0.1, ]
topSig_genes <- rownames(res_sig)
mat_sig_0.1 <- vsd_mat[topSig_genes, ]
###

mat_sig <- mat_sig_0.1 

### Make the Heatmap
# Scale each gene (row) to z-scores
mat_scaled_sig <- t(scale(t(mat_sig)))

# Optional sample annotation
annotation_col <- data.frame(
  Condition = colData(vsd)$condition)
rownames(annotation_col) <- colnames(mat_scaled_sig)

# Heatmap
pheatmap(
  mat_scaled_sig,
  annotation_col = annotation_col,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 6
)
```

# Make a heatmap with adjusted p-values < 0.05 and color code for the manuscript
```{r}
## Keep genes with adjusted p-value < 0.1
res_sig <- res[!is.na(res$padj) & res$padj < 0.05, ]
topSig_genes <- rownames(res_sig)
mat_sig_0.05 <- vsd_mat[topSig_genes, ]
###

mat_sig <- mat_sig_0.05 
mat_sig

### Make the Heatmap
# Scale each gene (row) to z-scores
mat_scaled_sig <- t(scale(t(mat_sig)))

# Optional sample annotation
annotation_col <- data.frame(
  Condition = colData(vsd)$condition)
rownames(annotation_col) <- colnames(mat_scaled_sig)

annotation_colors <- list(
  Succumb = c(
    Yes = "purple",
    No = "black"
  )
)

colnames(annotation_col)[colnames(annotation_col) == "Condition"] <- "Succumb"

# Heatmap
pheatmap(
  mat_scaled_sig,
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 8
)
```

  
 ###  Make a PCA from top genes with adjusted p-value < 0.05
```{r}
library(ggplot2)
library(ggrepel)

# PCA expects samples in rows and genes in columns
pca <- prcomp(t(mat_sig_0.05), scale. = TRUE)


# Percent variance explained
percentVar <- pca$sdev^2 / sum(pca$sdev^2)
# Create plotting dataframe
pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  condition = colData(vsd)$condition,
  sample = colnames(vsd)
)


ggplot(pca_df, aes(PC1, PC2, color = condition)) +
  geom_point(size = 4) +
  xlab(paste0("PC1: ", round(percentVar[1] * 100, 1), "%")) +
  ylab(paste0("PC2: ", round(percentVar[2] * 100, 1), "%")) +
  geom_text_repel(aes(label = sample)) +
  scale_color_manual(values = c("Yes" = "purple",
                                "No" = "black")) +
  labs(color = "Succumb") +
  theme_bw()

#ggplot(pca_df, aes(PC1, PC2, color = condition)) +
 # geom_point(size = 4) +
#  xlab(paste0("PC1: ", round(percentVar[1] * 100, 1), "%")) +
 # ylab(paste0("PC2: ", round(percentVar[2] * 100, 1), "%")) +
  #geom_text_repel(aes(label = sample))  +
  #theme_bw()  
```

### Preparing Data for GSEA. 
##  Rename first column so it matches "gene_id" in annotation file
```{r}
names(resOrdered)[1]<- "gene_id" 
```

#### Make a list of all the genes ranked by the p-value and logfold change, that can be brought into GSEA 
## example aa <- within(resOrdered, z <- x + y - 2)
```{r}
DGE_Rank <-  within(resOrdered, rank <- sign(log2FoldChange) * -log10(pvalue))
DGE_Rank 
```

#subset the results so only Gene Name and rank
```{r}
DGErank = subset(DGE_Rank, select = c(rank) )
DGErank
```

#remove any genes that don't ahve a name
```{r}
DGErank_withName <- na.omit(DGErank)
DGErank_withName
dim(DGErank_withName)
```

## Write out the file with the ranked list for GSEA - will use the GUI interface.
```{r}
write.table(as.data.frame(DGErank_withName), file="DeDup_DGErankName_OnlyPubescent.rnk", quote=FALSE, row.names=TRUE, sep = "\t")
```

####  Get the normalized expression DATA, incase need it.
```{r}
nt <- normTransform(dds_1f) # defaults to log2(x+1)
head(assay(nt))
```

# make it a new dataframe
```{r}
NormTransExp<-assay(nt)
summary(NormTransExp)
head(NormTransExp)
```

# Write the Normalized Transfored Data 
```{r}
write.table(as.data.frame(NormTransExp), file="Dedup_NormTransExp_OnlyPubescent.txt", quote=FALSE, row.names=TRUE, sep = "\t")  
```


#### GSEA was conducted on the GUI interface. 
  ##The enriched pathways were organized in to the Hallmarks.csv file for plotting.
  ## We used AI to troubleshoot this section of code.
```{r}
library(ggplot2)
Hallmark <- read.csv("Hallmarks.csv")

ggplot(Hallmark,
       aes(x = EnrichmentScore,
           y = reorder(Name, EnrichmentScore),
           size = GeneCount,
           color = NOM_pvalue)) +

  geom_point() +

  #facet_grid(. ~ Direction) +

  scale_color_gradient(
    low = "red",
    high = "steelblue",
    trans = "log10",
    name = "p-value",
    limits = c(0.00001, 0.25),
    guide = guide_colorbar(reverse = TRUE)
  ) +

  scale_size(range = c(3,10)) +

  labs(
    title = "GSEA Hallmarks Pathways",
    x = "Enrichment Score",
    y = NULL
  ) +

  theme_bw() +

  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_rect(fill = "grey85"),
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 11)
  )

```
